Scaling rates of true polar wander in convecting planets and moons

### 2016-08-08 tpw_rate.tex
Reference chapter.

@@

The term on the right-hand side of Equation~\eqref{eq:disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
This part of the moment of inertia may also be parameterized using a viscoelastic Love number approach:
\begin{equation} 
\mathbf{E} = \left[ \delta(t) + k^L(t) \right] * \mathbf{C}
\end{equation}
where $k^L$ is an internal loading Love number representing the surface deflection due to density anomalies, 
and $C_{ij}$ is the moment of inertia due solely to the internal load.
Frequently the simplification is made that the timescale of the surface response is quick compared to the true polar wander timescale, and we may use the fluid limit geoid kernels \citep[e.g.][]{richards1984geoid}:  
\begin{equation}
E_{ij} = (1+k^L_f) C_{ij}.
\end{equation}
An alternative to the Love number formalism is to calculate to surface deflections directly using mantle convection simulations with a true free surface boundary condition.
A recent implementation of a free surface boundary condition in the CIG-sponsored mantle convection software \texttt{ASPECT} \citep{rose2016free}
permits more general treatments of mantle rheology.

@@

\begin{equation}
\frac{\partial T}{\partial t} + \mathbf{u} \cdot \nabla T = \kappa \nabla^2 T
\label{eq:energy}
\end{equation}
where $\mathbf{u}$ is the velocity, $P$ is the pressure, and $T$ is the temperature.
The vector $\mathbf{g}$ is the gravitational acceleration, 
which defined in terms of a gravitational potential $V$ by $\mathbf{g} = -\nabla V$. 
The gravitational potential obeys 
\begin{equation}
\nabla^2 V = 4 \pi G \rho.
\end{equation}
where $G$ is the gravitational constant. 
For the purposes of scaling, we assume that the magnitude of the gravitational acceleration, $g_0$, is the approximately constant (as is the case for Earth's mantle).
In addition we use the simple equation of state
\begin{equation}
\rho = \rho_0 \left( 1 - \alpha (T-T_0) \right).
\label{eq:eos}
\end{equation}
to define the density.
The remaining parameters are defined in Table~\ref{tab:parameters}.
Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure
(in the latter case the boundary conditions on $P$ must be modified).
The centrifugal term, although the largest of the terms neglected in typical mantle convection models,
is generally small compared to gravitational forces (at least for Earthlike parameters),
and likely does not have a strong influence on the style of convection.
However, this term is critical for determining the size of the rotational bulge which must move relative to the mantle,
and so we must include it to establish the connection between the linear and angular momentum equations.

@@

For thermal convection, the convective part of the moment of inertia $E_{ij}$ is directly related to the degree-two part of the temperature field (see Appendix~\ref{appendix:moments}).
Therefore, an estimate for one specifies the other.
We can expand the temperature structure of the convecting planet with a set of orthonormal basis functions $R_n Y_{lm}$, 
where $Y_{lm}$ are spherical harmonics, $R_n$ are some set of orthogonal radial polynomials, and
$T_{nlm}$ are the coefficents for the expansion which have been normalized by $\Delta T$:
\begin{equation} 
T( r , \theta, \phi, t )  = \Delta T {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi).
\label{eq:T_series}
\end{equation}
Inserting the temperature expansion~\eqref{eq:T_series} into Equation~\eqref{eq:temperature_fluctuations} results in 
a prefactor of the nondimensional number $\Gamma = \alpha \Delta T$.
Orthogonality of the basis functions for the expansion means that the integral for $C_{ij}$ picks out degree-two 
spherical harmonics in the lateral dimensions, and only the lowest few radial functions $R_n(r)$.
Therefore, of the entire temperature spectrum, only a few of the modes matter for TPW.
We want to estimate the power in those few modes, which we denote by $T_{\text{degree-two}}$ (see Appendix~\ref{appendix:moments} for more detail): 
\begin{equation}
\Lambda_{31} \sim \Gamma T_{\text{degree-two}}.
\end{equation}
The temperature field has been normalized by $\Delta T$ and thus goes between zero and one, therefore the expansion in $T_{lmn}$ is constrained by

@@

In the limit that the eigengap becomes zero, the rotation of the pinciple axes can be arbitrary. 
This essentially corresponds to the hypothesized ``inertial interchange true polar wander'' \citep{kirschvink1997evidence}, where 
the mismatch angle is $90^\circ$. However, the eigengap does not need to be zero for there to be large displacements polar wander, 
and if it does go to zero, the wander does not need to be $90^\circ$.

@@

The theory of TPW response for the case of elastic lithospheres has been developed in, among other places, 
\citet{matsuyama2006rotational}, \citet{creveling2012mechanisms}, and \citet{chan2014time}.
The formalism developed in Section~\ref{sec:scaling} can still be applied to this case, 
though the response to internal variations in the moment of inertia becomes more limited 
(and potentially richer, as in the oscillitory motions suggested by \citet{creveling2012mechanisms}).

@@

\section*{Acknowledgments}
This work was supported by National Science Foundation grant EAR-1246670. We thank Yanick Ricard for careful and constructive comments that greatly improved the paper.
We thank the Computational Infrastructure for Geodynamics (geodynamics.org) which is funded by the National Science Foundation under award NSF-0949446.

\section*{Bibliography}
\bibliographystyle{elsarticle-harv} 
\bibliography{tpw_rate}

\appendix

@@

There is a connection between the moment of inertia of a rotating object and the degree-two density structure.
The moment of inertia tensor may be written in index notation
\begin{equation}
I_{ij} = \int_V \rho \left( r_q r_q \delta_{ij} - r_i r_j \right) dV
\label{eq:inertia}
\end{equation}
where $\mathbf{r}$ is the Eulerian coordinate, $\rho$ is the density, and $V$ is the volume of the material.  
It is useful to enter the principal axes of the moment of inertia:
\begin{equation}
\mathbf{I} = \mathbf{1} \begin{bmatrix}
\lambda_1 \\ \lambda_2 \\ \lambda_3
\end{bmatrix} = 
\mathbf{1} \begin{bmatrix}
\int_V \rho (y^2+z^z) dV\\
\int_V \rho (x^2+z^2) dV\\
\int_V \rho (x^2+y^2) dV
\end{bmatrix}
\end{equation}
where $\mathbf{1}$ is the identity matrix, and $\lambda_1$, $\lambda_2$, and $\lambda_3$ are the principal moments.
From Equation~\eqref{eq:polar_wander_rate} we see that the important quantities are the differences between the 
principal moments, $(\lambda_3-\lambda_1)$, $(\lambda_3-\lambda_2)$ and $(\lambda_2-\lambda_1)$.
These quantities may be rewritten in terms of degree-two real sperical harmonics \citep[e.g.][]{dahlen1999theoretical}.
The relevant (fully normalized) harmonics are, in Cartesian coordinates:
\begin{equation}
\begin{aligned}
Y_{20} &= \frac{1}{4} \sqrt{\frac{ 5}{\pi}} \frac{ 2 z^2 - x^2 - y^2}{r^2} \\ 
Y_{22} &= \frac{1}{4} \sqrt{\frac{15}{\pi}} \frac{ x^2 - y^2}{r^2}.
\end{aligned}
\end{equation}
Solving for $( \lambda_i - \lambda_j )$ in terms of these harmonics, we find
\ifdetail

It is useful to define unnormalized verions of the spherical harmonics as intermediate products:

@@

Up to the normalization constants, these expressions are idential to multipole expansions, 

@@

Dividing Equation~\eqref{eq:degree_two_moments_of_temperature} by $I_0$ and nodimensionalizing 

@@

>>>>

%%BEGINCLIP

@@

The term on the right-hand side of Equation~\eqref{eq:disequilibrium} represents the moment of inertia due to internal density anomalies as well as the surface deflections due to them.
This part of the moment of inertia may also be parameterized using a viscoelastic Love number approach:
\begin{equation} 
\mathbf{E} = \left[ \delta(t) + k^L(t) \right] * \mathbf{C}
\end{equation}
where $k^L$ is an internal loading Love number representing the surface deflection due to density anomalies, 
and $C_{ij}$ is the moment of inertia due solely to the internal load.
Frequently the simplification is made that the timescale of the surface response is quick compared to the true polar wander timescale, and we may use the fluid limit geoid kernels \citep[e.g.][]{richards1984geoid}:  
\begin{equation}
E_{ij} = (1+k^L_f) C_{ij}.
\end{equation}
An alternative to the Love number formalism is to calculate to surface deflections directly using mantle convection simulations with a true free surface boundary condition.
In Chapter~\ref{chap:free_surface} we describe the implementation of a free surface boundary condition in the CIG-sponsored mantle convection software \texttt{ASPECT}.
permits more general treatments of mantle rheology.

@@

\begin{equation}
\frac{\partial T}{\partial t} + \mathbf{u} \cdot \nabla T = \kappa \nabla^2 T
\label{eq:energy}
\end{equation}
where $\mathbf{u}$ is the velocity, $P$ is the pressure, and $T$ is the temperature.
The vector $\mathbf{g}$ is the gravitational acceleration, 
which defined in terms of a gravitational potential $V$ by $\mathbf{g} = -\nabla V$. 
The gravitational potential obeys 
\begin{equation}
\nabla^2 V = 4 \pi G \rho.
\end{equation}
where $G$ is the gravitational constant. 
For the purposes of scaling, we assume that the magnitude of the gravitational acceleration, $g_0$, is the approximately constant (as is the case for Earth's mantle).
In addition we use the simple equation of state
\begin{equation}
\rho = \rho_0 \left( 1 - \alpha (T-T_0) \right).
\label{eq:eos}
\end{equation}
to define the density.
The remaining parameters are defined in Table~\ref{tab:parameters}.
Note that here we retain the centrifugal term, which is normally either neglected or absorbed into a modified pressure
(in the latter case the boundary conditions on $P$ must be modified).
The centrifugal term, although the largest of the terms neglected in typical mantle convection models,
is generally small compared to gravitational forces (at least for Earth-like parameters),
and likely does not have a strong influence on the style of convection.
However, this term is critical for determining the size of the rotational bulge which must move relative to the mantle,
and so we must include it to establish the connection between the linear and angular momentum equations.

@@

For thermal convection, the convective part of the moment of inertia $E_{ij}$ is directly related to the degree-two part of the temperature field (see Appendix~\ref{appendix:moments}).
Therefore, an estimate for one specifies the other.
We can expand the temperature structure of the convecting planet with a set of orthonormal basis functions $R_n Y_{lm}$, 
where $Y_{lm}$ are spherical harmonics, $R_n$ are some set of orthogonal radial polynomials, and
$T_{nlm}$ are the coefficients for the expansion which have been normalized by $\Delta T$:
\begin{equation} 
T( r , \theta, \phi, t )  = \Delta T {\displaystyle \sum_{n=0}^\infty \sum_{l=0}^\infty \sum_{m=-l}^{l} } T_{lmn}(t) R_n(r) Y_{lm} (\theta , \phi).
\label{eq:T_series}
\end{equation}
Inserting the temperature expansion~\eqref{eq:T_series} into Equation~\eqref{eq:temperature_fluctuations} results in 
a prefactor of the nondimensional number $\Gamma = \alpha \Delta T$.
Orthogonality of the basis functions for the expansion means that the integral for $C_{ij}$ picks out degree-two 
spherical harmonics in the lateral dimensions, and only the lowest few radial functions $R_n(r)$.
Therefore, of the entire temperature spectrum, only a few of the modes matter for TPW.
We want to estimate the power in those few modes, which we denote by $T_{\text{degree-two}}$ (see Appendix~\ref{appendix:moments} for more detail): 
\begin{equation}
\Lambda_{31} \sim \Gamma T_{\text{degree-two}}.
\end{equation}
The temperature field has been normalized by $\Delta T$ and thus goes between zero and one, therefore the expansion in $T_{lmn}$ is constrained by

@@

In the limit that the eigengap becomes zero, the rotation of the principle axes can be arbitrary. 
This essentially corresponds to the hypothesized ``inertial interchange true polar wander'' \citep{kirschvink1997evidence}, where 
the mismatch angle is $90^\circ$. However, the eigengap does not need to be zero for there to be large displacements polar wander, 
and if it does go to zero, the wander does not need to be $90^\circ$.

@@

The theory of TPW response for the case of elastic lithospheres has been developed in, among other places, 
\citet{matsuyama2006rotational}, \citet{creveling2012mechanisms}, and \citet{chan2014time}.
The formalism developed in Section~\ref{sec:scaling} can still be applied to this case, 
though the response to internal variations in the moment of inertia becomes more limited 
(and potentially richer, as in the oscillatory motions suggested by \citet{creveling2012mechanisms}).

@@

%\section*{Acknowledgments}
%This work was supported by National Science Foundation grant EAR-1246670. We thank Yanick Ricard for careful and constructive comments that greatly improved the paper.
%We thank the Computational Infrastructure for Geodynamics (geodynamics.org) which is funded by the National Science Foundation under award NSF-0949446.

%\section*{Bibliography}
%\bibliographystyle{elsarticle-harv} 
%\bibliography{tpw_rate}

\begin{subappendices}

@@

There is a connection between the moment of inertia of a rotating object and the degree-two density structure.
The moment of inertia tensor may be written in index notation
\begin{equation}
I_{ij} = \int_V \rho \left( r_q r_q \delta_{ij} - r_i r_j \right) dV
\label{eq:inertia}
\end{equation}
where $\mathbf{r}$ is the Eulerian coordinate, $\rho$ is the density, and $V$ is the volume of the material.  
It is useful to enter the principal axes of the moment of inertia:
\begin{equation}
\mathbf{I} = \mathbf{1} \begin{bmatrix}
\lambda_1 \\ \lambda_2 \\ \lambda_3
\end{bmatrix} = 
\mathbf{1} \begin{bmatrix}
\int_V \rho (y^2+z^z) dV\\
\int_V \rho (x^2+z^2) dV\\
\int_V \rho (x^2+y^2) dV
\end{bmatrix}
\end{equation}
where $\mathbf{1}$ is the identity matrix, and $\lambda_1$, $\lambda_2$, and $\lambda_3$ are the principal moments.
From Equation~\eqref{eq:polar_wander_rate} we see that the important quantities are the differences between the 
principal moments, $(\lambda_3-\lambda_1)$, $(\lambda_3-\lambda_2)$ and $(\lambda_2-\lambda_1)$.
These quantities may be rewritten in terms of degree-two real spherical harmonics \citep[e.g.][]{dahlen1999theoretical}.
The relevant (fully normalized) harmonics are, in Cartesian coordinates:
\begin{equation}
\begin{aligned}
Y_{20} &= \frac{1}{4} \sqrt{\frac{ 5}{\pi}} \frac{ 2 z^2 - x^2 - y^2}{r^2} \\ 
Y_{22} &= \frac{1}{4} \sqrt{\frac{15}{\pi}} \frac{ x^2 - y^2}{r^2}.
\end{aligned}
\end{equation}
Solving for $( \lambda_i - \lambda_j )$ in terms of these harmonics, we find
\ifdetail

It is useful to define unnormalized versions of the spherical harmonics as intermediate products:

@@

Up to the normalization constants, these expressions are identical to multipole expansions, 

@@

Dividing Equation~\eqref{eq:degree_two_moments_of_temperature} by $I_0$ and nondimensionalizing 

@@

\section{Derivation of the polar wander rate equation}
\label{appendix:scalar_liouville}

Here is the tedious algebra for deriving Equation~\eqref{eq:polar_wander_rate}.  
We begin with the scalar equation Equation~\eqref{eq:polar_wander_rate_tensor}.
Let the prefactors $\frac{1}{(C-A)T_1}$ be denoted by $A$:

\begin{equation}
\begin{aligned}
\dot{\mitbf{\omega}} \cdot \dot{\mitbf{\omega}} = \dot{\Theta}^2  &= 
A^2 \left[ \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
 - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2.
\right]\\
\end{aligned}
\end{equation}
Since this is a scalar equation, we can arbitrarily choose a coordinate system in which to evaluate it.
It is most convenient to choose the principal axes of $\mathbf{E}$, where
\begin{equation}
\mathbf{E} = 
\begin{bmatrix}
\lambda_1 & 0 & 0 \\
0 & \lambda_2 & 0 \\
0 & 0 & \lambda_3 \\
\end{bmatrix}
\end{equation}
\begin{equation}
\mitbf{\omega} = 
\begin{bmatrix}
\sin{\theta} \cos{\phi} \\
\sin{\theta} \sin{\phi} \\
\cos{\theta} \\
\end{bmatrix}.
\end{equation}
As a warmup, consider the case where $\phi = 0$, then  
\begin{equation}
\mitbf{\omega} = 
\begin{bmatrix}
\sin{\theta}  \\
0 \\
\cos{\theta} \\
\end{bmatrix}
\end{equation}
We can evaluate some of the terms of the scalar equation for $\dot{\Theta}^2$:
\begin{equation}
\mathbf{E}\cdot \mitbf{\omega} = 
\begin{bmatrix}
\lambda_1 \sin{\theta} \\
0 \\
\lambda_3 \cos{\theta}
\end{bmatrix}
\end{equation}

\begin{equation}
\left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 =  
\lambda_1^2 \sin^4{\theta} + \lambda_3^2 \cos^4{\theta} + 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta}
\end{equation}
Therefore, plugging these into the scalar equation for $\dot{\mitbf{\omega}}^2$, we find:
\begin{equation}
\begin{aligned}
\frac{\dot{\Theta}}{A^2}  &= 
 \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
 - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2  \\
 &= \lambda_1^2 \sin^2{\theta} + \lambda_3^2 \cos^2{\theta}
 - \lambda_1^2 \sin^4{\theta} - \lambda_3^2 \cos^4{\theta} - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
 &= \lambda_1^2 \sin^2{\theta} (1 - \sin^2{\theta}) + \lambda_3^2 \cos^2{\theta} (1 - \cos^2{\theta})
  - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
 &= \lambda_1^2 \sin^2{\theta} \cos^2{\theta} + \lambda_3^2 \cos^2{\theta} \sin^2{\theta}
  - 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \\
 &= (\lambda_1^2 + \lambda_3^2 - 2 \lambda_1 \lambda_3) \sin^2{\theta} \cos^2{\theta} \\
 &= \frac{1}{2} (\lambda_3 - \lambda_1)^2 \sin{2 \theta} \\
\end{aligned}
\end{equation}
which is equivalent to Equation~\eqref{eq:simple_milankovitch}.

Now consider the more general case where $\phi \ne 0$.
Again, we evaluate some of the terms of the scalar equation for $\dot{\mitbf{\omega}}^2$:
\begin{equation}
\mathbf{E}\cdot \mitbf{\omega} = 
\begin{bmatrix}
\lambda_1 \sin{\theta} \cos{\phi}\\
\lambda_2 \sin{\theta} \sin{\phi} \\
\lambda_3 \cos{\theta}
\end{bmatrix}
\end{equation}

\begin{equation}
\begin{aligned}
\left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2 =  
&\lambda_1^2 \sin^4{\theta} \cos^4{\phi} + 
\lambda_2^2 \sin^4{\theta} \sin^4{\phi} +
\lambda_3^2 \cos^4{\theta} \\
&+ 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
&+ 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
&+ 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi} \\ 
\end{aligned}
\end{equation}
Again, plugging these into the scalar equation for $\dot{\Theta}^2$, we find:
\begin{equation}
\begin{aligned}
\frac{\dot{\Theta}^2}{A^2}  &= 
 \left(\mathbf{E}\cdot \mitbf{\omega}\right)^2
 - \left(\mitbf{\omega} \cdot \mathbf{E} \cdot \mitbf{\omega} \right)^2  \\
&=
\lambda_1^2 \sin^2{\theta} \cos^2{\phi} + \lambda_2^2 \sin^2{\theta} \sin^2{\phi} + \lambda_3^2 \cos^2{\theta} \\
&-\lambda_1^2 \sin^4{\theta} \cos^4{\phi} - 
\lambda_2^2 \sin^4{\theta} \sin^4{\phi} -
\lambda_3^2 \cos^4{\theta} \\
&- 2 \lambda_1 \lambda_2 \sin^4{\theta} \cos^2{\phi}\sin^2{\phi} \\ 
&- 2 \lambda_1 \lambda_3 \sin^2{\theta} \cos^2{\theta} \cos^2{\phi} \\
&- 2 \lambda_2 \lambda_3 \sin^2{\theta} \cos^2{\theta} \sin^2{\phi}. 
\end{aligned}
\end{equation}
After a lot of simplification, we get
\begin{equation}
\begin{aligned}
\frac{\dot{\Theta}^2}{A^2}  = 
&\frac{1}{4}\sin^2{2\theta} \left[( \lambda_3 - \lambda_1)^2 \cos^2{\phi} + (\lambda_3 - \lambda_2)^2 \sin^2{\phi} \right] +\\
&\frac{1}{4}\sin^4{\theta} \sin^2{2\phi}\left( \lambda_2 - \lambda_1 \right)^2 \\
\end{aligned}
\end{equation}
which is Equation~\eqref{eq:polar_wander_rate}.

\end{subappendices}

%%ENDCLIP
